Opening files and getting started.
========================================================

Obviously, as this workshop progresses it might be impractical to use these outlines since we will be creating files in an ongoing manner.

This workshop is to help you see how to start a project (using pre-defined data) and how to open a variety of files you might come into contact with.

If you've created a GitHub account and forked (and cloned) this repository you will see that you have a number of folders[1].  The repository already has a few folders filled.  In particular, take a look at `data`.  

Folder | Contents | File Type | open file function | package
---------- | -------- | ----- | ----------- | ---------
input  | `compiled.pollen.RData` | R data file | `load` | {base}
input\Ecoregions_NA | `na_regns` | Shapefile | `readOGR` | {rgdal}
input\PalEONSites | `PaleonSites` | Shapefile  | `readOGR` | {rgdal}
input\USHCN | `9641C_201112_F52.pcp` | fixed width text file | `read_fwf` | {base}
input\USHCN | `ushcn-stations.csv`   | comma separated values | `read.csv` | {base}

So here we have three kinds of data (generally).  We have spatial data for North American ecozones, we have spatio-temporal data for precipiation (over the last 110 years) for the United States and we have spatio-temporal pollen data across North America for the last ~21000 years.

Let's start by opening up the pollen data.  First, lets look at where the points are located by opening the metadata file:

```{r LoadUSHCN, echo = TRUE, message = FALSE}
#  This file is located in the folder 'docs', let's move the working directory down to the base directory:

setwd('..')

#  Now open the station data.  By default R converts strings to a factor, but this 
#  is annoying behaviour, so we change the behaviour, setting 'stringsAsFactors' to FALSE
stations <- read.csv('input/USHCN/ushcn-stations.csv', stringsAsFactors=FALSE)

head(stations)

```

Lets plot out the station locations:

```{r fig.width=6, fig.height=6}
plot(Lat ~ Long, data = stations)
```

Now lets see how these fit into ecozones.
```{r}
library(rgdal)

file <- list.files('../input/Ecoregions_NA', full.names = TRUE)
file <- file[regexpr('shp$', file)>0]
ecozones <- readOGR(dsn = file, layer = 'na_regns')

# reproject ecozones
ecozones_ll <- spTransform(ecozones, CRSobj = CRS('+proj=longlat +ellps=WGS84'))

plot(ecozones_ll, xlim=c(-130, -50), ylim=c(25, 50))
points(Lat ~ Long, data = stations, pch=19, cex = 0.5, col = 2)
axis(1); axis(2)
```

So there is the distribution of points within the continental United States.  Easy peasy!

Next we're going to do some basic analysis on the data.


[1]: See [Getting started with GitHub](http://downwithtime.wordpress.com/2013/09/13/writing-and-collaborating-on-github-a-primer-for-paleoecologists/) for an example.
